Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-06-17 article-models-tt.tex
version esystems sent to VV and RG

\section{The algebraic theory of contexts, families and terms}\label{tt}
In this section we give a description of dependent type theory before type
constructors. Apart from contexts, families and terms -- which provide for the
core of the language of dependent logic -- the basic ingredients
of this theory will be the operations of extension, weakening and substitution
and the identity terms. An empty context (and empty families) are also included
in the theory. The resulting theory can be seen as a manifestation of the 
structure underlying dependent type theory.

We will
formulate the theory of contexts, families and terms in such a way that contexts aren't defined
to be lists of variable declarations. The variable-free (or name-free) approach 
we take here is rather different than those appearing in 
\cite{hofmann1995extensional,TheBook} but it has appeared in the work of Coquand
and in \cite{Dybjer1996}.
The main reason we don't let variable declarations in is that we don't see them 
in the internal models either. This way we also set out to a more algebraic 
approach of type theory and higher category theory. Thirdly, we will not have to
be burdened with superficial comments about variables being bounded or not, or 
fresh or free or not occuring at all.

In our treatment, a context can be the empty context or it can be a binary
planar tree of which `the leaves are (families of) contexts'. 
The judgmental equality relation on contexts is an equivalence relation which 
expresses that binary planar
trees of contexts are judgmentally equal if their leaves are, taking only
(the isomorphism class of) 
the order of the leaves into account \emph{and not the actual shape of the three}.
The intuition behind this equivalence relation is indeed that unstructured
(i.e.~unbracketed) lists such as the lists of variable declarations which
usually appear in type theoretical syntax, may be regarded as contexts.

Besides contexts, families and terms there will also be a notion of `type in
a context', which one may assume.
Only a family of contexts over a context $\Gamma$ is eligible to
be a type in the context $\Gamma$ and we have the intuition that `being a type'
expresses the property of `being an atomic or irreducible family of contexts'.
We will not axiomatize that every context is either judgmentally equal to the
empty context or judgmentally equal to the extension of a context by a type,
although it is certainly worth investigating the class of contexts which fall
under this category. However, we will primarily investigate families of
contexts and leave discussion on the possibility of a typing judgment until the
end of this section. With families of contexts being the principal things of
study, we have added a bit more generality to the theory, compared to the
theory of categories with families presented in \cite{Dybjer1996}. The reason
for this added generality is twofold:
\begin{enumerate}
\item We want to be very sure that we can interpret all the rules of type theory
in a straightforward manner in any context.
\item We do want do end up with a categorical structure on the contexts, derived
from the theory by requiring that a morphism from $\Gamma$ to $\Delta$ is a
term of the weakening of $\Delta$ by $\Gamma$ in context $\Gamma$. Thus, we
must be able to lift the context $\Delta$ to become a family of contexts over
$\Gamma$ and it is not reasonable to expect that this results in a type.
\end{enumerate}

@@

\subsection{The operations of the theory of contexts, families and terms}
In this subsection we introduce the notions of empty context,
extension, weakening, substitution and identity terms. These are all viewed as
operations of the theory of contexts, families and terms.

Generally, 
each operation should
have an action on contexts, on families and on terms at the `appropriate level'
and should be compatible with each of the operations we introduce here. In
that will become more practical after the theory of contexts, families and terms
is formulated, this means that when we will write down a type theory that
interprets itself in \autoref{ttmodels}, each of the notions of empty context,
extension, weakening, substitution and identity terms are going to be internalized
as terms of the theory and all terms are going to be compatible with these.
That vague idea will be our guiding light.

@@

Apart from the empty context, there will indeed be an empty family and a term
of the empty family (which will be deemed unique). But extension will
not have an explicit action `at the level of families' since extension
is used to define higher families and the action would have to be trivial
anyway. The empty context and extension provide the essentials of the rest of
the theory in the sense that higher families are families over extensions,
contexts are families over the empty context and the `action on families' of
any operation applied to families over the empty context is the same as the
original operation. Thus, when we have introduced the operations of weakening
and substitution we will see that they also have an `action on the terms of 
contexts'. We will state the relevant judgmental equality rules for the empty
context in this section rather than in the next, since they are required to
state that the action on families of an operation is compatible with the
original operation.

@@

\paragraph{Associativity of extension}
\label{comp-ee}
The inference rules asserting that extension is compatible with itself assert
that contexts are unstructured lists of type declarations. This rule is
unavoidable if we want that for a family $A$ in context $\Gamma$, a family over
$A$ is the same thing as a family over $\ctxext{\Gamma}{A}$. 
\begin{align}
%& \inference
%  { \jfam{\Gamma}{A}
%    \jfam{{\Gamma}{A}}{P}
%    }
%  { \jctxeq{\ctxext{{\Gamma}{A}}{P}}{\ctxext{\Gamma}{{A}{P}}}
%    }
%  \\
& \inference
  { \jfam{{\Gamma}{A}}{P}
    \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jfameq{\Gamma}{\ctxext{A}{{P}{Q}}}{\ctxext{{A}{P}}{Q}}
    }
  \label{comp-ee-c}
\end{align}
Note that we may take the point of view that extension distributes over
extensions. If we want to extend an extended family $\ctxext{P}{Q}$ over
$\ctxext{\Gamma}{A}$ by the family $A$, the result is the extension of (1)
the extension of $P$ by $A$ and (2) the extension of $Q$ by $A$. The latter is
simply $Q$ since extension acts trivially at the level of families (and terms).

@@

\subsubsection{Weakening}
\label{weakening}
When $A$ is a family in context $\Gamma$, the operation of weakening by $A$
takes a family $B$ in context $\Gamma$ and provides a family $\ctxwk{A}{B}$
in context $\ctxext{\Gamma}{A}$. The context family $\ctxwk{A}{B}$ can be seen
as the constant family over $\ctxext{\Gamma}{A}$ with value $B$. This idea will
be axiomatized in the cancellation property of weakening and substitution in
\autoref{cancellation-ws}. In \autoref{morphisms} we will take the terms 
$\unfold{\jhom{\Gamma}{A}{B}{f}}$ to be the morphisms of families from $A$ to 
$B$. These will be at the heart of the categorical structure of the theory.

The weakening operation acts on three levels: on contexts, on families and
on terms. The `action on contexts' of weakening is the action we described
above: it takes a family $B$ over $\Gamma$ to a family $\ctxwk{A}{B}$ over
$\ctxext{\Gamma}{A}$; the `action on families' of weakening takes a family
$Q$ over $\ctxext{\Gamma}{B}$ to a family $\ctxwk[\famsym]{A}{Q}$ over
$\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}$; the `action on terms' of weakening takes
a term $g$ of $Q$ to a term $\ctxwk[\tmsym]{A}{g}$ of $\ctxwk[\famsym]{A}{Q}$.
\begin{align}
& \inference
  { \jfam{\Gamma}{A}
    \jfam{\Gamma}{B}
    }
  { \jfam{{\Gamma}{A}}{\ctxwk{A}{B}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      \jfameq{\Gamma}{B}{B'}
      }
    { \jfameq{{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A'}{B'}}
      }
    \\
& \inference
  { \jfam{\Gamma}{A}
    \jfam{{\Gamma}{B}}{Q}
    }
  { \jfam{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk[\famsym]{A}{Q}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      \jfameq{{\Gamma}{B}}{Q}{Q'}
      }
    { \jfameq
        {{{\Gamma}{A}}{\ctxwk{A}{B}}}
        {\ctxwk[\famsym]{A}{Q}}
        {\ctxwk[\famsym]{A'}{Q'}}
      }
    \\
%& \inference
%  { \jterm{\Gamma}{B}{y}
%    }
%  { \jterm{{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}
%    }
%& & \inference
%    { \jtermeq{\Gamma}{B}{y}{y'}
%      }
%    { \jtermeq
%        {{\Gamma}{A}}
%        {\ctxwk{A}{B}}
%        {\ctxwk{A}{y}}
%        {\ctxwk{A'}{y'}}
%      }
%    \\
& \inference
  { \jfam{\Gamma}{A}
    \jterm{{\Gamma}{B}}{Q}{g}
    }
  { \jterm{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk[\famsym]{A}{Q}}{\ctxwk[\tmsym]{A}{g}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      \jtermeq{{\Gamma}{B}}{Q}{g}{g'}
      }
    { \jtermeq
        {{{\Gamma}{A}}{\ctxwk{A}{B}}}
        {\ctxwk[\famsym]{A}{Q}}
        {\ctxwk[\tmsym]{A}{g}}
        {\ctxwk[\tmsym]{A'}{g'}}
      }
\end{align}
The following inference rules express that when the empty context or family is
weakened, the result is the empty family.
\label{comp-w0}
\begin{align}
%& \inference
%  { \jctx{\Gamma}
%    }
%  { \jfameq{\Gamma}{\ctxwk{\Gamma}{\emptyc}}{\emptyf}
%    }
%  \\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{{\Gamma}{A}}{\ctxwk{A}{\emptyf}}{\emptyf}
    }
  \label{comp-w0-c}\\
& \inference
  { \jfam{\Gamma}{A}
    \jfam{\Gamma}{B}
    }
  { \jfameq
    {{{\Gamma}{A}}{\ctxwk{A}{B}}}
    {\ctxwk[\famsym]{A}{\emptyf}}
    {\emptyf}
    }
  \label{comp-w0-f}
\end{align}
Note that we do not have to axiomatize the action of weakening
by $A$ on the term of the empty family, because the result will be in the empty 
family and every term in the empty family is judgmentally equal to the term 
$\emptytm$.

@@

\subsubsection{Substitution}
\label{substitution}
Given a family $P$ over $A$ and a term $x$ of $A$, substitution gives a way to
consider the \emph{fiber $\subst{x}{P}$ of $P$ at $x$}. As was the case with
weakening, the substitution operation comes in three parts: the `action on
contexts' of substitution is the operation just described; the `action on
families' of substitution takes a family $Q$ over $\ctxext{{\Gamma}{A}}{P}$
to a family $\subst[\famsym]{x}{Q}$ over $\ctxext{\Gamma}{\subst{x}{P}}$; the
`action on terms' of substitution takes a term $g$ of $Q$ over
$\ctxext{{\Gamma}{A}}{P}$ to a term $\subst[\tmsym]{x}{g}$ of 
$\subst[\famsym]{x}{Q}$.
\begin{align}
& \inference
  { \jterm{\Gamma}{A}{x}
    \jfam{{\Gamma}{A}}{P}
    }
  { \jfam{\Gamma}{\subst{x}{P}}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{x'}
      \jfameq{{\Gamma}{A}}{P}{P'}
      }
    { \jfameq{\Gamma}{\subst{x}{P}}{\subst{x'}{P'}}
      }
    \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jfam{{\Gamma}{\subst{x}{P}}}{\subst[\famsym]{x}{Q}}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{x'}
      \jfameq{{{\Gamma}{A}}{P}}{Q}{Q'}
      }
    { \jfameq{{\Gamma}{\subst{x}{P}}}{\subst[\famsym]{x}{Q}}{\subst[\famsym]{x'}{Q'}}
      }
    \\
%& \inference
%  { \jterm{{\Gamma}{A}}{P}{f}
%    }
%  { \jterm{\Gamma}{\subst{x}{P}}{\subst{x}{f}}
%    }
%& & \inference
%    { \jtermeq{{\Gamma}{A}}{P}{f}{f'}
%      }
%    { \jtermeq{\Gamma}{\subst{x}{P}}{\subst{x}{f}}{\subst{x'}{f'}}
%      }
%    \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{{{\Gamma}{A}}{P}}{Q}{g}
    }
  { \jterm{{\Gamma}{\subst{x}{P}}}{\subst[\famsym]{x}{Q}}{\subst[\tmsym]{x}{g}}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{x'}
      \jtermeq{{{\Gamma}{A}}{P}}{Q}{g}{g'}
      }
    { \jtermeq
        {{\Gamma}{\subst{x}{P}}}
        {\subst[\famsym]{x}{Q}}
        {\subst[\tmsym]{x}{g}}
        {\subst[\tmsym]{x'}{g'}}
      }
\end{align}
The following inference rules establish that the fibers of the empty family are 
the empty families:
\label{comp-s0}
\begin{align}
& \inference
  { \jterm{\Gamma}{A}{x}
    }
  { \jfameq{\Gamma}{\subst{x}{\emptyf}}{\emptyf}
    }
  \label{comp-s0-c}
  \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jfam{{\Gamma}{A}}{P}
    }
  { \jfameq
      {{\Gamma}{\subst{x}{P}}}
      {\subst{x}{\emptyf}}
      {\emptyf}
    }
  \label{comp-s0-f}
\end{align}
Note again that we do not have to axiomatize the action of substitution by $x$ on the term
of the empty family, because the result will be in the empty family and every
term in the empty family is judgmentally equal to the term $\emptytm$.

@@

\subsubsection{Identity terms}
\label{identityterms}
Last, but not least, we introduce the identity terms. In combination with
weakening and substitution, identity terms will allow us to speak of projections,
of a variant of extension for terms and of fiber inclusions. The following
inference rules are required to be valid:
\begin{align}
& \inference
  { \jfam{\Gamma}{A}
    }
  { \unfoldall{\jhom{\Gamma}{A}{A}{\idtm{A}}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      }
    { \unfoldall{\jhomeq{\Gamma}{A}{A}{\idtm{A}}{\idtm{A'}}}
      }
\end{align}

@@

\subsubsection{Substituting by the term of the empty family}
\label{comp-0s}
The following rules assert that substituting by the term $\jterm{\Gamma}{\emptyf}{\emptytm}$
leaves everything unchanged.
\begin{align}
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{\subst{\emptytm}{A}}{A}
    }
  \label{comp-0s-c}
  \\
%& \inference
%  { \jterm{\Gamma}{A}{x}
%    }
%  { \jtermeq{\Gamma}{A}{\subst{\emptytm}{x}}{x}
%    }
%  \\
& \inference
  { \jfam{{\Gamma}{A}}{P}
    }
  { \jfameq{{\Gamma}{A}}{\subst{\emptytm}{P}}{P}
    }
  \label{comp-0s-f}
  \\
& \inference
  { \jterm{{\Gamma}{A}}{P}{f}
    }
  { \jtermeq{{\Gamma}{A}}{P}{\subst{\emptytm}{f}}{f}
    }.
  \label{comp-0s-t}
\end{align}

>>>>

\section{The generalized algebraic theory of contexts, families and terms (the E-system)}\label{tt}
In this section we give a description of dependent type theory before type
constructors. Apart from contexts, families and terms -- which provide for the
core of the language of dependent logic -- the basic ingredients
of this theory will be the operations of extension, empty families, weakening and substitution
and the identity terms. The resulting theory can be seen as a manifestation of the 
structure underlying dependent type theory.

We will
formulate the theory of contexts, families and terms in such a way that contexts aren't defined
to be lists of variable declarations. The variable-free (a.k.a.~\emph{name-free}) approach 
we take here is rather different than those appearing in 
\cite{hofmann1995extensional,TheBook} but it has appeared in the work of Coquand
and in \cite{Dybjer1996}.
The main reason we don't let variable declarations in is that we don't see them 
in the internal models either. This way we also set out to a more algebraic 
approach of type theory and higher category theory. Thirdly, we will not have to
be burdened with superficial comments about variables being bounded or not, or 
fresh or free or not occuring at all.

In our treatment, a contexts can be seen as binary
planar trees of which the leaves are (families of) contexts. 
The judgmental equality relation on contexts is an equivalence relation which 
expresses that binary planar
trees of contexts are judgmentally equal if their leaves are, taking only
(the isomorphism class of) 
the order of the leaves into account \emph{and not the actual shape of the three}.
The intuition behind this equivalence relation is indeed that unstructured
(i.e.~unbracketed) lists such as the lists of variable declarations which
usually appear in type theoretical syntax, may be regarded as contexts.

Besides contexts, families and terms there will also be a notion of `type in
a context', which one may assume.
A context is eligible to be a \emph{closed type} and over a context, a family
may be a type. We have the intuition that `being a type'
expresses the property of `being atomic or irreducible'. With a notion of types,
one has the means to require a stratification by which every context is a finite
extension of types starting at a closed type, and every family is likewise a
finite extension of types. The idea behind this is to recover the theory of
\autoref{Garner2014} {\color{red}(and possibly B-systems?)}. We stress that
we will not include axioms asserting such a stratification in the E-system
and we leave discussion on the possibility of a typing judgment until the
end of this section.

With families of contexts being the principal things of
study, we have added a bit more generality to the theory, compared to the
theory of categories with families presented in \cite{Dybjer1996}. Here are
a couple of reasons for this extra generality:
\begin{enumerate}
\item We want to be very sure that we can interpret all the rules of type theory
in a straightforward manner in any context.
\item We do want do end up with a categorical structure on the contexts, derived
from the theory by requiring that a morphism from $\Gamma$ to $\Delta$ is a
term of the weakening of $\Delta$ by $\Gamma$ in context $\Gamma$. Thus, we
must be able to lift the context $\Delta$ to become a family of contexts over
$\Gamma$ and it is not reasonable to expect that this results in a type.
\item It is very easy in this setting to assert that an operation has an action
on very high levels of families, i.e.~when $F$ is an operation taking things in
context $\Gamma$ to things in context $\Delta$, then it takes a family $A$ in
context $\Gamma$ to $F(A)$ in context $\Delta$; a family $P$ in context
$\ctxext{\Gamma}{A}$ to a family $F(P)$ in context $\ctxext{\Delta}{(F(A))}$,
and so on. The reason it becomes simple is that we don't have to explain what
`and so on' means.
\end{enumerate}

@@

\subsection{The operations of the theory of contexts, families and terms}
In this subsection we introduce the notions of
extension, empty families, 
weakening by families, weakening by contexts, 
substitution and identity terms. The weakenings and substitution are both
operations of the theory of contexts, families and terms and therefore they will
have three levels of actions: on contexts (at the appropriate level), families
and terms.

\begin{comment}
Generally, 
each operation should
have an action on contexts, on families and on terms at the `appropriate level'
and should be compatible with each of the operations we introduce here. In
that will become more practical after the theory of contexts, families and terms
is formulated, this means that when we will write down a type theory that
interprets itself in \autoref{ttmodels}, each of the notions of
extension, empty family, weakening, substitution and identity terms are going to be internalized
as terms of the theory and all terms are going to be compatible with these.
That vague idea will be our guiding light.

@@

Apart from the empty context, there will indeed be an empty family and a term
of the empty family (which will be deemed unique). But extension will
not have an explicit action `at the level of families' since extension
is used to define higher families and the action would have to be trivial
anyway. The empty context and extension provide the essentials of the rest of
the theory in the sense that higher families are families over extensions,
contexts are families over the empty context and the `action on families' of
any operation applied to families over the empty context is the same as the
original operation. Thus, when we have introduced the operations of weakening
and substitution we will see that they also have an `action on the terms of 
contexts'. We will state the relevant judgmental equality rules for the empty
context in this section rather than in the next, since they are required to
state that the action on families of an operation is compatible with the
original operation.
\end{comment}

We note upfront that both context extension and family extension can also be
seen as operations of contexts, families and terms, even though we will not
introduce them as such. The reason is that extension is used to define what it
means to be a family over a family over a context (namely, a family over an
extended context). As such, it will have a trivial action on families and terms
which will not need to be introduced. We will come back to this point in
\autoref{extension}.

@@

\paragraph{Associativity of extension}
\label{comp-ee}
The inference rules asserting that extension is compatible with itself assert
that contexts are unstructured lists of type declarations. This rule is
unavoidable if we want that for a family $A$ in context $\Gamma$, a family over
$A$ is the same thing as a family over $\ctxext{\Gamma}{A}$. 
\begin{align}
& \inference
  { \jfam{\Gamma}{A}
    \jfam{{\Gamma}{A}}{P}
    }
  { \jctxeq{\ctxext{{\Gamma}{A}}{P}}{\ctxext{\Gamma}{{A}{P}}}
    }
  \\
& \inference
  { \jfam{{\Gamma}{A}}{P}
    \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jfameq{\Gamma}{\ctxext{A}{{P}{Q}}}{\ctxext{{A}{P}}{Q}}
    }
  \label{comp-ee-c}
\end{align}
Note that we may take the point of view that extension distributes over
extensions. If we want to extend an extended family $\ctxext{P}{Q}$ over
$\ctxext{\Gamma}{A}$ by the family $A$, the result is the extension of (1)
the extension of $P$ by $A$ and (2) the extension of $Q$ by $A$. The latter is
simply $Q$ since extension acts trivially at the level of families (and terms).

@@

\subsubsection{Weakening by a family}
\label{weakening}
When $A$ is a family in context $\Gamma$, the operation of weakening by $A$
takes a family $B$ in context $\Gamma$ and provides a family $\ctxwk{A}{B}$
in context $\ctxext{\Gamma}{A}$. The context family $\ctxwk{A}{B}$ can be seen
as the constant family over $\ctxext{\Gamma}{A}$ with value $B$. This idea will
be axiomatized in the cancellation property of weakening and substitution in
\autoref{cancellation-ws}. In \autoref{morphisms} we will take the terms 
$\unfold{\jhom{\Gamma}{A}{B}{f}}$ to be the morphisms of families from $A$ to 
$B$. These will be at the heart of the categorical structure of the theory.

The weakening operation acts on three levels: on contexts, on families and
on terms. The `action on contexts' of weakening is the action we described
above: it takes a family $B$ over $\Gamma$ to a family $\ctxwk{A}{B}$ over
$\ctxext{\Gamma}{A}$; the `action on families' of weakening takes a family
$Q$ over $\ctxext{\Gamma}{B}$ to a family $\ctxwk[\famsym]{A}{Q}$ over
$\ctxext{{\Gamma}{A}}{\ctxwk{A}{B}}$; the `action on terms' of weakening takes
a term $g$ of $Q$ to a term $\ctxwk[\tmsym]{A}{g}$ of $\ctxwk[\famsym]{A}{Q}$.
\begin{align}
& \inference
  { \jfam{\Gamma}{A}
    \jfam{\Gamma}{B}
    }
  { \jfam{{\Gamma}{A}}{\ctxwk{A}{B}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      \jfameq{\Gamma}{B}{B'}
      }
    { \jfameq{{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A'}{B'}}
      }
    \\
& \inference
  { \jfam{\Gamma}{A}
    \jfam{{\Gamma}{B}}{Q}
    }
  { \jfam{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk[\famsym]{A}{Q}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      \jfameq{{\Gamma}{B}}{Q}{Q'}
      }
    { \jfameq
        {{{\Gamma}{A}}{\ctxwk{A}{B}}}
        {\ctxwk[\famsym]{A}{Q}}
        {\ctxwk[\famsym]{A'}{Q'}}
      }
    \\
%& \inference
%  { \jterm{\Gamma}{B}{y}
%    }
%  { \jterm{{\Gamma}{A}}{\ctxwk{A}{B}}{\ctxwk{A}{y}}
%    }
%& & \inference
%    { \jtermeq{\Gamma}{B}{y}{y'}
%      }
%    { \jtermeq
%        {{\Gamma}{A}}
%        {\ctxwk{A}{B}}
%        {\ctxwk{A}{y}}
%        {\ctxwk{A'}{y'}}
%      }
%    \\
& \inference
  { \jfam{\Gamma}{A}
    \jterm{{\Gamma}{B}}{Q}{g}
    }
  { \jterm{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\ctxwk[\famsym]{A}{Q}}{\ctxwk[\tmsym]{A}{g}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      \jtermeq{{\Gamma}{B}}{Q}{g}{g'}
      }
    { \jtermeq
        {{{\Gamma}{A}}{\ctxwk{A}{B}}}
        {\ctxwk[\famsym]{A}{Q}}
        {\ctxwk[\tmsym]{A}{g}}
        {\ctxwk[\tmsym]{A'}{g'}}
      }
\end{align}
The following inference rules express that when the empty family is
weakened, the result is the empty family.
\label{comp-w0}
\begin{align}
%& \inference
%  { \jctx{\Gamma}
%    }
%  { \jfameq{\Gamma}{\ctxwk{\Gamma}{\emptyc}}{\emptyf}
%    }
%  \\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{{\Gamma}{A}}{\ctxwk{A}{\emptyf}}{\emptyf}
    }
  \label{comp-w0-c}\\
& \inference
  { \jfam{\Gamma}{A}
    \jfam{\Gamma}{B}
    }
  { \jfameq
    {{{\Gamma}{A}}{\ctxwk{A}{B}}}
    {\ctxwk[\famsym]{A}{\emptyf}}
    {\emptyf}
    }
  \label{comp-w0-f}
\end{align}
\begin{comment}
Note that we do not have to axiomatize the action of weakening
by $A$ on the term of the empty family, because the result will be in the empty 
family and every term in the empty family is judgmentally equal to the term 
$\emptytm$.
\end{comment}

@@

\subsubsection{Weakening by a context}
Since we have not required that there is an empty context with the property that
a context is the same thing as a family in the empty context, we will have to
explicitly introduce weakening by contexts. Since it will allways be clear
whether something is weakened by a context or by a family, we shall use the
same notation for weakening by contexts as for weakening by families. We shall
have to repeat all the compatibility rules for weakening by families for
weakening by contexts; there is no way around it. It is
actually this sort of inconvenience that has originally led us to include an
empty context and be done with it.

In the following inference rules we assume that we have a context $\Gamma$ and
that $\jctxeq{\Gamma}{\Gamma'}$ in the rules where $\Gamma'$ makes an appearence.
\begin{align}
& \inference
  { \jctx{\Delta}
    }
  { \jfam{\Gamma}{\ctxwk{\Gamma}{\Delta}}
    }
& & \inference
  { \jctxeq{\Delta}{\Delta'}
    }
  { \jfameq{\Gamma}{\ctxwk{\Gamma}{\Delta}}{\ctxwk{\Gamma'}{\Delta'}}
    }
  \\
& \inference
  { \jfam{\Delta}{B}
    }
  { \jfam{\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}{\ctxwk{\Gamma}{B}}
    }
& & \inference
  { \jfameq{\Delta}{B}{B'}
    }
  { \jfameq
      {\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}
      {\ctxwk{\Gamma}{B}}
      {\ctxwk{\Gamma'}{B'}}
    }
  \\
& \inference
  { \jterm{\Delta}{B}{y}
    }
  { \jterm
      {\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}
      {\ctxwk{\Gamma}{B}}
      {\ctxwk{\Gamma}{y}}
    }
& & \inference
  { \jtermeq{\Delta}{B}{y}{y'}
    }
  { \jtermeq
      {\ctxext{\Gamma}{\ctxwk{\Gamma}{\Delta}}}
      {\ctxwk{\Gamma}{B}}
      {\ctxwk{\Gamma}{y}}
      {\ctxwk{\Gamma}{y'}}
    }
\end{align}
Note that we do not employ three different notations for the three different
actions of weakening by a context because there can be no confusion about which
of the three actions is meant. If there were an empty context we should have
distinguished the three as we did for weakening by a family.

We shall have similar compatibility rules for weakening by a context as we did
for weakening by a family.

{\color{red}To do. It will look exactly the same as for weakening by families}

\subsubsection{Substitution}
\label{substitution}
Given a family $P$ over $A$ and a term $x$ of $A$, substitution gives a way to
consider the \emph{fiber $\subst{x}{P}$ of $P$ at $x$}. As was the case with
weakening, the substitution operation comes in three parts: the `action on
contexts' of substitution is the operation just described; the `action on
families' of substitution takes a family $Q$ over $\ctxext{{\Gamma}{A}}{P}$
to a family $\subst[\famsym]{x}{Q}$ over $\ctxext{\Gamma}{\subst{x}{P}}$; the
`action on terms' of substitution takes a term $g$ of $Q$ over
$\ctxext{{\Gamma}{A}}{P}$ to a term $\subst[\tmsym]{x}{g}$ of 
$\subst[\famsym]{x}{Q}$.
\begin{align}
& \inference
  { \jterm{\Gamma}{A}{x}
    \jfam{{\Gamma}{A}}{P}
    }
  { \jfam{\Gamma}{\subst{x}{P}}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{x'}
      \jfameq{{\Gamma}{A}}{P}{P'}
      }
    { \jfameq{\Gamma}{\subst{x}{P}}{\subst{x'}{P'}}
      }
    \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jfam{{{\Gamma}{A}}{P}}{Q}
    }
  { \jfam{{\Gamma}{\subst{x}{P}}}{\subst[\famsym]{x}{Q}}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{x'}
      \jfameq{{{\Gamma}{A}}{P}}{Q}{Q'}
      }
    { \jfameq{{\Gamma}{\subst{x}{P}}}{\subst[\famsym]{x}{Q}}{\subst[\famsym]{x'}{Q'}}
      }
    \\
%& \inference
%  { \jterm{{\Gamma}{A}}{P}{f}
%    }
%  { \jterm{\Gamma}{\subst{x}{P}}{\subst{x}{f}}
%    }
%& & \inference
%    { \jtermeq{{\Gamma}{A}}{P}{f}{f'}
%      }
%    { \jtermeq{\Gamma}{\subst{x}{P}}{\subst{x}{f}}{\subst{x'}{f'}}
%      }
%    \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jterm{{{\Gamma}{A}}{P}}{Q}{g}
    }
  { \jterm{{\Gamma}{\subst{x}{P}}}{\subst[\famsym]{x}{Q}}{\subst[\tmsym]{x}{g}}
    }
& & \inference
    { \jtermeq{\Gamma}{A}{x}{x'}
      \jtermeq{{{\Gamma}{A}}{P}}{Q}{g}{g'}
      }
    { \jtermeq
        {{\Gamma}{\subst{x}{P}}}
        {\subst[\famsym]{x}{Q}}
        {\subst[\tmsym]{x}{g}}
        {\subst[\tmsym]{x'}{g'}}
      }
\end{align}
The following inference rules establish that the fibers of the empty family are 
the empty families:
\label{comp-s0}
\begin{align}
& \inference
  { \jterm{\Gamma}{A}{x}
    }
  { \jfameq{\Gamma}{\subst{x}{\emptyf}}{\emptyf}
    }
  \label{comp-s0-c}
  \\
& \inference
  { \jterm{\Gamma}{A}{x}
    \jfam{{\Gamma}{A}}{P}
    }
  { \jfameq
      {{\Gamma}{\subst{x}{P}}}
      {\subst{x}{\emptyf}}
      {\emptyf}
    }
  \label{comp-s0-f}
\end{align}
\begin{comment}
Note again that we do not have to axiomatize the action of substitution by $x$ on the term
of the empty family, because the result will be in the empty family and every
term in the empty family is judgmentally equal to the term $\emptytm$.
\end{comment}

@@

\subsubsection{Identity terms}
\label{identityterms}
Last, but not least, we introduce the identity terms. In combination with
weakening and substitution, identity terms will allow us to speak of projections,
of a variant of extension for terms and of fiber inclusions.
The following
inference rules are required to be valid:
\begin{align}
& \inference
  { \jfam{\Gamma}{A}
    }
  { \unfoldall{\jhom{\Gamma}{A}{A}{\idtm{A}}}
    }
& & \inference
    { \jfameq{\Gamma}{A}{A'}
      }
    { \unfoldall{\jhomeq{\Gamma}{A}{A}{\idtm{A}}{\idtm{A'}}}
      }
\end{align}
We stress again
that because we have not required there to be an empty context, we need to
explicitly call for identity terms $\idtm{\Gamma}$ of $\ctxwk{\Gamma}{\Gamma}$
in context $\Gamma$ for each context $\Gamma$. Thus, we will also require
the inference rule
\begin{align}
& \inference
  { \jctx{\Gamma}
    }
  { \jterm{\Gamma}{\ctxwk{\Gamma}{\Gamma}}{\idtm{\Gamma}}
    }
& &
  \inference
  { \jctxeq{\Gamma}{\Gamma'}
    }
  { \jtermeq{\Gamma}{\ctxwk{\Gamma}{\Gamma}}{\idtm{\Gamma}}{\idtm{\Gamma'}}
    }
\end{align}
For the following inference rules, there should also be versions of identity
terms at contexts.

@@

\begin{comment}
\subsubsection{Substituting by the term of the empty family}
\label{comp-0s}
The following rules assert that substituting by the term $\jterm{\Gamma}{\emptyf}{\emptytm}$
leaves everything unchanged.
\begin{align}
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{\subst{\emptytm}{A}}{A}
    }
  \label{comp-0s-c}
  \\
%& \inference
%  { \jterm{\Gamma}{A}{x}
%    }
%  { \jtermeq{\Gamma}{A}{\subst{\emptytm}{x}}{x}
%    }
%  \\
& \inference
  { \jfam{{\Gamma}{A}}{P}
    }
  { \jfameq{{\Gamma}{A}}{\subst{\emptytm}{P}}{P}
    }
  \label{comp-0s-f}
  \\
& \inference
  { \jterm{{\Gamma}{A}}{P}{f}
    }
  { \jtermeq{{\Gamma}{A}}{P}{\subst{\emptytm}{f}}{f}
    }.
  \label{comp-0s-t}
\end{align}
\end{comment}
